Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-12-12 papers/contact/paper.tex
move contact figures around Tested everything on kipu.
more changes for resubmission of contact Tested everything on kipu.

Here we will briefly describe how the contact value of the correlation
function has been used in two of these recent papers introducing SAFT-based
classical density functionals.  For simplicity, we
will use our own notation to describe the work of these authors.  In
his paper presenting a density functional based on the PCP-SAFT
equation of state\cite{gross2009density}, Gross introduces the chain
free energy in SAFT as
\begin{align}
  \frac{A_\textit{chain}}{kT} &= -(m-1) \int n(\rr) \left(\ln\left(n^A(\rr)
  g_\sigma^A(\rr) \right)-1\right) d\rr \label{eq:Achain}
\end{align}
where $n^A(\rr)$ is a weighted density defined in Equation~\ref{eq:nA}
and $g_\sigma^A(\rr)$ the local value of the correlation function at
contact, which we define in Equation~\ref{eq:gA}.  In
Section~\ref{sec:gross} we describe Gross's approximation for this
function. In a paper describing a classical
density functional for inhomogeneous associating
fluids\cite{yu2002fmt-dft-inhomogeneous-associating}, Yu and Wu define
the association free energy as
\begin{align}
  \frac{A_\textit{assoc}}{kT} &= \sum_i \int \zeta(\rr)n_0(\rr) \left(\ln X_i(\rr) - \frac12
  X_i(\rr) + \frac12\right) d\rr \label{eq:Aassoc} \\
  X_i(\rr) &= \frac{1}{1 + \sum_j \frac{n_0(\rr) g_\sigma^{S}(\rr)}{\zeta(\rr)}
                                 X_j(\rr)\kappa_{ij} \left(e^{\beta
                                   \epsilon_{ij}}-1\right)}
  \label{eq:X}
\end{align}
where $X_i(\rr)$ is the fraction of interaction sites of type $i$ at
position $\rr$ that are unoccupied, $n_0(\rr)$ is a weighted density
defined in Equation~\ref{eq:n0}, $\zeta(\rr)$ is a nonlocal measure of
the density gradient defined in Equation~\ref{eq:zeta}, and
$g_\sigma^S(\rr)$ is a form of the correlation
function at contact, which we define in Equation~\ref{eq:gS}.  In
Section~\ref{sec:yuwu} we present the approximation to
$g_\sigma^S(\rr)$ introduced by Yu and Wu.
Given these differing approaches, it seems valuable to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

@@

In order to motivate our derivation of the correlation function at
contact for the \emph{inhomogeneous} hard-sphere fluid, we begin by
deriving the well-known formula for $g_\sigma$ for the
\emph{homogeneous} fluid that comes from the Carnahan-Starling free
energy.  The contact value of the correlation function density can be
found by using the contact-value theorem, which states that the
pressure on any hard surface is determined by the density at contact:
\begin{align}
  p(\rr_c) &= k_BT n(\rr_c)
\end{align}
where $\rr_c$ is a position of a sphere that would be in contact with
the hard surface, $n(\rr_c)$ is the density at this point $\rr_c$, and $p(\rr_c)$ is the pressure that the spheres
exert on the surface at the same point.  This pressure is
understood as the ratio of force to an infinitesimal element of
\emph{solvent accessible surface} (SAS) area.
%
In the homogeneous fluid, the contact-value theorem implies that
\begin{align}
  p_{\textit{SAS}} &= k_BT n g_\sigma
\end{align}
where $p_{SAS}$ is the pressure on a hard sphere's solvent accessible surface.
This pressure can be
readily computed from the dependence of the Carnahan-Starling free
energy on hard sphere radius:
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{\pi}{6} \sigma^3 n$ is the filling fraction.
We compute the pressure using the total force with
respect to a changing radius of all the spheres.
To find the pressure, we divide this force by $4\pi \sigma^2$, which
is the SAS area of a single hard sphere, illustrated in Figure~\ref{fig:nA}.
Finally, we divide by $N$ to account for the total area of all the
spheres in the fluid.
\begin{align}
  p_{\textit{SAS}} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
  &= \frac{1}{N 4\pi \sigma^2} \frac12 \frac{dA_{HS}}{d\sigma} \\
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we thus find the well-known
correlation function evaluated at contact.
\begin{align}
  g_\sigma &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Extending this derivation to the inhomogeneous fluid requires that we
find the pressure felt by the surface of particular spheres.

@@

We will begin our derivation of the locally averaged correlation
function with the asymmetric definition of $g_\sigma^A(\rr)$ given in
Equation~\ref{eq:gA}, which is averaged over contacts in which one of
the two spheres is located at position~$\rr$.  This correlation
function is related to the contact density averaged over the solvent accessible surface
of a sphere located at~$\rr$, and can thus be determined by finding
the pressure on that surface.  We find this pressure from the change in
free energy resulting from an infinitesimal expansion of spheres
located at position~$\rr$.  From this pressure, we derive a
formula for the correlation function $g_\sigma^A(\rr)$ as was done in
the previous section:
\begin{align}
  p_{\textit{SAS}}(\mathbf{r}) &= \frac{1}{n(\rr) 4\pi \sigma^2}\frac12 \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
%  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 4\pi \sigma^2}\frac12 \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr)% &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 4\pi \sigma^2}\frac12 \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where $\sigma(\mathbf{r})$ is the diameter of spheres located at position $\rr$.
Details regarding the evaluation of the functional derivative $\frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
are discussed in the Appendix.
Equation~\ref{eq:g-A-exact} is an exact expression for
$g_\sigma^A(\rr)$ as defined in Equation~\ref{eq:gA}.  However, since
we do not know the exact hard-sphere free energy, we approximate
$A_{HS}$ using FMT.  Details of evaluating the functional derivative
in Equation~\ref{eq:g-A-exact} using FMT are given in the appendix.
The equation for $g_\sigma^A$ requires finding convolutions of local
derivatives of the free energy, making this formulation
computationally somewhat more expensive than the free energy itself.

@@

We now address the symmetrically averaged correlation function, which
is defined in Equation~\ref{eq:gS}.  This corresponds to the
correlation function averaged for spheres \emph{touching at a given
  point}.  In this case, we conceptually would like to evaluate the
pressure felt by the surface of spheres where that surface is located
at point~$\rr$.  We can approximate this value by assuming that this
pressure will be simply related to the free energy density at
point~$\rr$.  Through a process similar to the previous derivations, this
leads to the expression
\begin{align}
  %% p_{HS}^S(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\partial
  %%   a_{HS}(\rr)}{\partial \sigma} \label{eq:p_{HS}^S}\\
%  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr)% &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ 4\pi \sigma^2}\frac12
  \frac{\partial \Phi(\rr)}{\partial \sigma} \label{eq:g-S}
\end{align}
where $\Phi(\rr) = \Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)$ is the
dimensionless free energy density.  This expression is an
approximation---unlike the analogous
Equation~\ref{eq:g-A-exact}---because it assumes that we have
available a local functional $\Phi(\rr)$ whose derivative provides the
pressure needed to compute $g_\sigma(\rr)$.  Equation~\ref{eq:g-S}
requires that we evaluate the derivatives of the fundamental measures
$n_\alpha(\rr)$ with respect to diameter, which leads us to
derivatives of the $\delta$~function, which we can simplify and
approximate using an assumption of a reasonably smooth density:
\begin{align}
  \frac{\partial n_2(\rr)}{\partial \sigma}
  &= \frac12 \int \delta'(\frac\sigma2 - |\rr-\rr'|) n(\rr')d\rr' \\
  &= \frac2{\sigma}n_2(\rr) - \frac12 \int \delta\left(\frac\sigma2 - |\rr-\rr'|\right)
  \frac{\rr-\rr'}{|\rr-\rr'|}\cdot\mathbf{\nabla}n(\rr')d\rr' \\
  &\approx \frac{2}{\sigma}n_2(\rr)
\end{align}
In the systems that we study, the density is \emph{not} reasonably
smooth, but we can state empirically making this approximation
nevertheless improves the predictions of our functional, while at the
same time reducing its computational cost by avoiding the need to
calculate any additional weighted densities or convolutions.

@@

We simulate the inhomogeneous hard sphere fluid at four hard-wall
interfaces.  The first and simplest is a flat hard wall.  We
then study two convex hard surfaces.  One is an excluded sphere with
diameter $2\sigma$, which corresponds to a ``test particle''
simulation with one of a hard sphere at the origin with diameter
$\sigma$.  The
second is an excluded sphere with diameter $6\sigma$,
which demonstrates behavior typical of mildly convex hard surfaces.
Finally, we study a concave surface given by a hard cavity in which
our fluid is free to move up to a diameter of $8\sigma$, which
demonstrates behavior typical of mildly concave surfaces.  In each
case, we performed a low-density (filling fraction 0.1) and high-density
(filling fraction 0.4) simulation.  We performed additional
computations over a wider range of curvatures and densities, but
chose these as typical examples.

@@

@@

We begin by presenting our low-density results, corresponding to a
filling fraction of 0.1, which are shown in
Figure~\ref{fig:low-density}.  At this low density,
the contact value of the correlation function in the bulk is only 1.3,
indicating that correlations are indeed small and that the fluid should be
relatively easy to model.  Indeed, the contact density at the hard
surface is only around 50\% higher than the bulk, and the FMT
predicted density is close to indistinguishable from the true
density for each of the four configurations.

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/inner-12-10}
%%   \caption{Density and correlation function around a hard sphere with
%%     an excluded diameter of $6\sigma$, with a bulk filling fraction of
%%     0.1.}
%%   \label{fig:inner-12-10}
%% \end{figure}

The $g_\sigma^A$ correlation function in each configuration is very
flat, with only small, smooth changes as the surface is approached.
Our functional $g_\sigma^A$ very closely matches the Monte Carlo
predictions in each case, while that of Gross consistently
underestimates the correlation at the interface by a significant margin.  We note that the theoretical curves
extend into the region from which the fluid is excluded.  This value
corresponds to the correlation function that would be observed in the
vanishingly unlikely scenario in which there was a sphere present at
that location.  Naturally, we are unable to observe this quantity in
our Monte Carlo simulations.

The $g_\sigma^S$ correlation function shows considerably more
structure, as well as additional variation due to the curvature of the
hard surface.  The symmetric correlation function is nonzero at
locations where spheres may touch, which for a convex hard surface
means that $g_\sigma^S$ may be nonzero in the volume in which hard
spheres are excluded.  In every configuration studied, the agreement
between the theoretical predictions and the Monte Carlo simulation in
each case is very poor in the region where there should be no contacts
at all.  Because $n_0$ is comparable to its bulk value in this region,
this means that these functionals predict a significant number of
contacts in the region where there should be none.  The correlation
function of Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}
and ours described in Section~\ref{sec:g-S} give similar results, with
slightly larger errors in our prediction.

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/outer-16-01}
%%   \caption{Density and correlation functions near the surface of a
%%     spherical cavity with diameter $16\sigma$ at a bulk filling
%%     fraction of 0.1.}
%%   \label{fig:outer-10}
%% \end{figure}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/test-particle-10}
%%   \caption{DFT and monte-carlo nA compared with averaged test particle data with bulk filling
%%     fraction of 0.1.}
%%   \label{fig:test-particle-10}
%% \end{figure}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/test-particle-40}
%%   \caption{DFT and monte-carlo nA compared with averaged test particle data with bulk filling
%%     fraction of 0.4.}
%%   \label{fig:test-particle-40}
%% \end{figure}

@@

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/walls-40}
%%   \caption{Density and correlation function at a hard wall with bulk
%%     filling fraction of 0.4.}
%%   \label{fig:walls-40}
%% \end{figure}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/inner-4-40}
%%   \caption{Density and correlation function around a hard sphere with
%%     diameter the same as those in the fluid, with a bulk filling
%%     fraction of 0.4.}
%%   \label{fig:inner-4-40}
%% \end{figure}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{figs/inner-12-40}
%%   \caption{Density and correlation function around a hard sphere with
%%     an excluded diameter of $6\sigma$, with a bulk filling fraction of
%%     0.4.}
%%   \label{fig:inner-12-40}
%% \end{figure}

\begin{figure*}
  \noindent\includegraphics[width=0.8\textwidth]{figs/low-density}
  %% \noindent
  %% \begin{tabular}{cc}
  %%   (a)\includegraphics[width=8cm]{figs/walls-10} &
  %%   (c)\includegraphics[width=8cm]{figs/inner-4-10}\\
  %%   (b)\includegraphics[width=8cm]{figs/inner-12-10} &
  %%   (d)\includegraphics[width=8cm]{figs/outer-16-01}
  %% \end{tabular}
  \caption{ (Color online) Density and correlation function in systems
    with a ``low density'' bulk filling fraction of 0.1.  The
    subplots each show a different system: (a) next to a flat hard wall, (b)
    around a hard sphere with an excluded diameter of $6\sigma$, (c)
    around a hard sphere with an excluded diameter of $2\sigma$, and
    (d) within a spherical cavity with an included diameter of $8\sigma$.
    In the top and middle panels of each subfigure are the
    asymmetrically averaged correlation function $g_\sigma^A$ (defined
    in Equation~\ref{eq:gA}) and the symmetrically averaged
    correlation function $g_\sigma^S$ (defined in
    Equation~\ref{eq:gS}).  The results of Monte Carlo, our
    functional, and one previously published
    functional\cite{gross2009density,
      yu2002fmt-dft-inhomogeneous-associating} are compared in each
    case.  The bottom panels show the density computed with
    Monte Carlo and with DFT.}
  \label{fig:low-density}
\end{figure*}

\begin{figure*}
  \noindent\includegraphics[width=0.8\textwidth]{figs/high-density}
  %% \noindent
  %% \begin{tabular}{cc}
  %%   (a)\includegraphics[width=8cm]{figs/walls-40} &
  %%   (c)\includegraphics[width=8cm]{figs/inner-4-40}\\
  %%   (b)\includegraphics[width=8cm]{figs/inner-12-40} &
  %%   (d)\includegraphics[width=8cm]{figs/outer-16-04}
  %% \end{tabular}
  \caption{ (Color online) Density and correlation function in systems
    with a ``high density'' bulk filling fraction of 0.4.  The
    subplots each show a different system: (a) next to a flat hard wall, (b)
    around a hard sphere with an excluded diameter of $6\sigma$, (c)
    around a hard sphere with an excluded diameter of $2\sigma$, and
    (d) within a spherical cavity with an included diameter of $8\sigma$.
    In the top and middle panels of each subfigure are the
    asymmetrically averaged correlation function $g_\sigma^A$ (defined
    in Equation~\ref{eq:gA}) and the symmetrically averaged
    correlation function $g_\sigma^S$ (defined in
    Equation~\ref{eq:gS}).  The results of Monte Carlo, our
    functional, and one previously published
    functional\cite{gross2009density,
      yu2002fmt-dft-inhomogeneous-associating} are compared in each
    case.  The bottom panels show the density computed with
    Monte Carlo and with DFT.}
  \label{fig:high-density}
\end{figure*}

At a higher density corresponding to a filling fraction of 0.4,
correlations are much stronger, with the bulk contact value of the
correlation function of 3.7, as seen in
Figure~\ref{fig:high-density}.  This results in larger
oscillations in the density at the hard surfaces, and correspondingly
more interesting behavior in the correlation function near the
interface.  The density predicted by the White Bear functional agrees
reasonably well with the simulation results, although not so well as
it did at lower density.  The discrepancies are largest in the case of
the spherical cavity (Figure~\ref{fig:high-density}d), in which the DFT considerably underestimates the
range of the density oscillations.

The asymmetric version of the correlation function once again displays
relatively smooth behavior with a few small
oscillations near the interface, and a somewhat elevated value within
a diameter of the hard surface, with the magnitude of this elevation
somewhat different in each configuration.  As was the case at low
density, our correlation function $g_\sigma^A$ matches very closely
the Monte Carlo data, reproducing quite well the structure near the
interface in each configuration, although in the spherical cavity there is a small, but significant
discrepancy, comparable to the discrepancy found in the density
itself.  In each case, the correlation of Gross dramatically
underestimates the correlation at the interface, at one extreme by
40\% in the case of the spherical cavity, and at the other extreme by 15\%
in the test-particle scenario (Figure~\ref{fig:high-density}c).

>>>>

Here we will briefly describe how the contact value of the correlation
function has been used in two of these recent papers introducing SAFT-based
classical density functionals.  For simplicity, we
will use our own notation to describe the work of these authors.  In
his paper presenting a density functional based on the PCP-SAFT
equation of state\cite{gross2009density}, Gross introduces the chain
free energy in SAFT as
\begin{align}
  \frac{A_\textit{chain}}{kT} &= -(m-1) \int n(\rr) \left(\ln\left(n^A(\rr)
  g_\sigma^A(\rr) \right)-1\right) d\rr \label{eq:Achain}
\end{align}
where $n^A(\rr)$ is a weighted density defined in Equation~\ref{eq:nA}
and $g_\sigma^A(\rr)$ is the local value of the correlation function at
contact, which we define in Equation~\ref{eq:gA}.  In
Section~\ref{sec:gross} we describe Gross's approximation for this
function. In a paper describing a classical
density functional for inhomogeneous associating
fluids\cite{yu2002fmt-dft-inhomogeneous-associating}, Yu and Wu define
the association free energy as
\begin{align}
  \frac{A_\textit{assoc}}{kT} &= \sum_i \int \zeta(\rr)n_0(\rr) \left(\ln X_i(\rr) - \frac12
  X_i(\rr) + \frac12\right) d\rr \label{eq:Aassoc} \\
  X_i(\rr) &= \frac{1}{1 + \sum_j \frac{n_0(\rr) g_\sigma^{S}(\rr)}{\zeta(\rr)}
                                 X_j(\rr)\kappa_{ij} \left(e^{\beta
                                   \epsilon_{ij}}-1\right)}
  \label{eq:X}
\end{align}
where $X_i(\rr)$ is the fraction of interaction sites of type $i$ at
position $\rr$ that are unoccupied, $n_0(\rr)$ is a weighted density
defined in Equation~\ref{eq:n0}, $\zeta(\rr)$ is a nonlocal measure of
the density gradient defined in Equation~\ref{eq:zeta}, and
$g_\sigma^S(\rr)$ is a form of the correlation
function at contact, which we define in Equation~\ref{eq:gS}.  In
Section~\ref{sec:yuwu} we present the approximation to
$g_\sigma^S(\rr)$ introduced by Yu and Wu.
Given these differing approaches, it seems valuable to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

@@

In order to motivate our derivation of the correlation function at
contact for the \emph{inhomogeneous} hard-sphere fluid, we begin by
deriving the well-known formula for $g_\sigma$ for the
\emph{homogeneous} fluid that comes from the Carnahan-Starling free
energy.  The contact value of the correlation function density can be
found by using the contact-value theorem, which states that the
pressure on any hard surface is determined by the density at contact:
\begin{align}
  p_{\textit{SAS}}(\rr_c) &= k_BT n(\rr_c)
\end{align}
where $\rr_c$ is a position of a sphere that would be in contact with
the hard surface, $n(\rr_c)$ is the density at this point $\rr_c$, and
$p_{\textit{SAS}}(\rr_c)$ is the pressure that the spheres
exert on the surface at the same point.  This pressure is
understood as the ratio of force to an infinitesimal element of
\emph{solvent accessible surface} (SAS) area.
%
In the homogeneous fluid, the contact-value theorem implies that
\begin{align}
  p_{\textit{SAS}} &= k_BT n g_\sigma
\end{align}
where $p_{SAS}$ is the pressure on a hard sphere's solvent accessible surface.
This pressure can be
readily computed from the dependence of the Carnahan-Starling free
energy on hard sphere radius:
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{\pi}{6} \sigma^3 n$ is the filling fraction.
We compute the pressure using the total force with
respect to a changing radius of all the spheres.
To find the pressure, we divide this force by $4\pi \sigma^2$, which
is the SAS area of a single hard sphere, illustrated in Figure~\ref{fig:nA}.
Finally, we divide by $N$ to account for the total area of all the
spheres in the fluid.
\begin{align}
  p_{\textit{SAS}} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
  &= \frac{1}{N 4\pi \sigma^2} \frac12 \frac{dA_{HS}}{d\sigma} \\
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we thus find the well-known
correlation function evaluated at contact.
\begin{align}
  g_\sigma &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Extending this derivation to the inhomogeneous fluid requires that we
find the pressure felt by the surface of particular spheres.

@@

We will begin our derivation of the locally averaged correlation
function with the asymmetric definition of $g_\sigma^A(\rr)$ given in
Equation~\ref{eq:gA}, which is averaged over contacts in which one of
the two spheres is located at position~$\rr$.  This correlation
function is related to the contact density averaged over the solvent accessible surface
of a sphere located at~$\rr$, and can thus be determined by finding
the pressure on that surface.  We find this pressure from the change in
free energy resulting from an infinitesimal expansion of spheres
located at position~$\rr$.  From this pressure, we derive a
formula for the correlation function $g_\sigma^A(\rr)$ as was done in
the previous section:
\begin{align}
  p_{\textit{SAS}}(\mathbf{r}) &= \frac{1}{n(\rr) 4\pi \sigma^2}\frac12 \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
%  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 4\pi \sigma^2}\frac12 \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr)% &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 4\pi \sigma^2}\frac12 \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where $\sigma(\mathbf{r})$ is the diameter of spheres located at position $\rr$.
Details regarding the evaluation of the functional derivative
$\frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
are discussed in the Appendix.
Equation~\ref{eq:g-A-exact} is an exact expression for
$g_\sigma^A(\rr)$ as defined in Equation~\ref{eq:gA}.  However, since
we do not know the exact hard-sphere free energy, we approximate
$A_{HS}$ using FMT.
The equation for $g_\sigma^A$ found using FMT requires finding
convolutions of local derivatives of the free energy, making this
formulation computationally somewhat more expensive than the free
energy itself.

@@

We now address the symmetrically averaged correlation function, which
is defined in Equation~\ref{eq:gS}.  This corresponds to the
correlation function averaged for spheres \emph{touching at a given
  point}.  In this case, we conceptually would like to evaluate the
pressure felt by the surface of spheres where that surface is located
at point~$\rr$.  We can approximate this value by assuming that this
pressure will be simply related to the free energy density at
point~$\rr$.  Through a process similar to the previous derivations, this
leads to the expression
\begin{align}
  %% p_{HS}^S(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\partial
  %%   a_{HS}(\rr)}{\partial \sigma} \label{eq:p_{HS}^S}\\
%  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr)% &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ 4\pi \sigma^2}\frac12
  \frac{\partial \Phi(\rr)}{\partial \sigma} \label{eq:g-S}
\end{align}
where $\Phi(\rr) = \Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)$ is the
dimensionless free energy density.  This expression is an
approximation---unlike the analogous
Equation~\ref{eq:g-A-exact}---because it assumes that we have
available a local functional $\Phi(\rr)$ whose derivative provides the
pressure needed to compute $g_\sigma(\rr)$.  Equation~\ref{eq:g-S}
requires that we evaluate the derivatives of the fundamental measures
$n_\alpha(\rr)$ with respect to diameter, which leads us to
derivatives of the $\delta$~function, which we can simplify and
approximate using an assumption of a reasonably smooth density:
\begin{align}
  \frac{\partial n_2(\rr)}{\partial \sigma}
  &= \frac12 \int \delta'(\frac\sigma2 - |\rr-\rr'|) n(\rr')d\rr' \\
  &= \frac2{\sigma}n_2(\rr) - \frac12 \int \delta\left(\frac\sigma2 - |\rr-\rr'|\right)
  \frac{\rr-\rr'}{|\rr-\rr'|}\cdot\mathbf{\nabla}n(\rr')d\rr' \\
  &\approx \frac{2}{\sigma}n_2(\rr)
\end{align}
In the systems that we study, the density is \emph{not} reasonably
smooth, but we can state empirically making this approximation
nevertheless improves the predictions of our functional $g_\sigma^S$, while at the
same time reducing its computational cost by avoiding the need to
calculate any additional weighted densities or convolutions.

@@

We simulate the inhomogeneous hard sphere fluid at four hard-wall
interfaces.  The first and simplest is a flat hard wall.  We
then study two convex hard surfaces.  One is an excluded sphere with
diameter $2\sigma$, which corresponds to a ``test particle''
simulation with one of a hard sphere at the origin with diameter
$\sigma$.  The
second is an excluded sphere with diameter $6\sigma$,
which demonstrates behavior typical of mildly convex hard surfaces.
Finally, we study a concave surface given by a hard cavity in which
our fluid is free to move up to a diameter of $16\sigma$, which
demonstrates behavior typical of mildly concave surfaces.  In each
case, we performed a low-density (filling fraction 0.1) and high-density
(filling fraction 0.4) simulation.  We performed additional
computations over a wider range of curvatures and densities, but
chose these as typical examples.

@@

\begin{figure*}
  \noindent\includegraphics[width=0.8\textwidth]{figs/low-density}
  %% \noindent
  %% \begin{tabular}{cc}
  %%   (a)\includegraphics[width=8cm]{figs/walls-10} &
  %%   (c)\includegraphics[width=8cm]{figs/inner-4-10}\\
  %%   (b)\includegraphics[width=8cm]{figs/inner-12-10} &
  %%   (d)\includegraphics[width=8cm]{figs/outer-16-01}
  %% \end{tabular}
  \caption{ (Color online) Density and correlation function in systems
    with a ``low density'' bulk filling fraction of 0.1.  The
    subplots each show a different system: (a) next to a flat hard wall, (b)
    around a hard sphere with an excluded diameter of $6\sigma$, (c)
    around a hard sphere with an excluded diameter of $2\sigma$, and
    (d) within a spherical cavity with an included diameter of $16\sigma$.
    In the top and middle panels of each subfigure respectively are the
    asymmetrically averaged correlation function $g_\sigma^A$ (defined
    in Equation~\ref{eq:gA}) and the symmetrically averaged
    correlation function $g_\sigma^S$ (defined in
    Equation~\ref{eq:gS}).  The results of Monte Carlo, our
    functional, and one previously published
    functional\cite{gross2009density,
      yu2002fmt-dft-inhomogeneous-associating} are compared in each
    case.  The bottom panels show the density computed with
    Monte Carlo and with DFT.}
  \label{fig:low-density}
\end{figure*}

@@

We begin by presenting our low-density results, corresponding to a
filling fraction of 0.1, which are shown in
Figure~\ref{fig:low-density}.  At this low density,
the contact value of the correlation function in the bulk is only 1.3,
indicating that correlations are indeed small and that the fluid should be
relatively easy to model.  Indeed, the contact density at the hard
surface is only around 50\% higher than the bulk, and the FMT
predicted density is close to indistinguishable from the true
density for each of the four configurations, as seen in the bottom
subpanel of each subfigure within Figure~\ref{fig:low-density}.

The $g_\sigma^A$ correlation function in each configuration (plotted
in the top panel of each subfigure within Figure~\ref{fig:low-density}) is
very flat, with only small, smooth changes as the surface is approached.
Our functional $g_\sigma^A$ very closely matches the Monte Carlo
predictions in each case, while that of Gross consistently
underestimates the correlation at the interface by a significant margin.  We note that the theoretical curves
extend into the region from which the fluid is excluded.  This value
corresponds to the correlation function that would be observed in the
vanishingly unlikely scenario in which there was a sphere present at
that location.  Naturally, we are unable to observe this quantity in
our Monte Carlo simulations.

The $g_\sigma^S$ correlation function (plotted
in the middle panels of Figure~\ref{fig:low-density}) shows considerably more
structure, as well as additional variation due to the curvature of the
hard surface.  The symmetric correlation function is nonzero at
locations where spheres may touch, which for a convex hard surface
means that $g_\sigma^S$ may be nonzero in the volume in which hard
spheres are excluded.  In every configuration studied, the agreement
between the theoretical predictions and the Monte Carlo simulation in
each case is very poor in the region where there should be no contacts
at all.  Because $n_0$ is comparable to its bulk value in this region,
this means that these functionals predict a significant number of
contacts in the region where there should be none.  The correlation
function of Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}
and ours described in Section~\ref{sec:g-S} give similar results, with
slightly larger errors in our prediction.

\begin{figure*}
  \noindent\includegraphics[width=0.8\textwidth]{figs/high-density}
  %% \noindent
  %% \begin{tabular}{cc}
  %%   (a)\includegraphics[width=8cm]{figs/walls-40} &
  %%   (c)\includegraphics[width=8cm]{figs/inner-4-40}\\
  %%   (b)\includegraphics[width=8cm]{figs/inner-12-40} &
  %%   (d)\includegraphics[width=8cm]{figs/outer-16-04}
  %% \end{tabular}
  \caption{ (Color online) Density and correlation function in systems
    with a ``high density'' bulk filling fraction of 0.4.  The
    subplots each show a different system: (a) next to a flat hard wall, (b)
    around a hard sphere with an excluded diameter of $6\sigma$, (c)
    around a hard sphere with an excluded diameter of $2\sigma$, and
    (d) within a spherical cavity with an included diameter of $16\sigma$.
    In the top and middle panels of each subfigure respectively are the
    asymmetrically averaged correlation function $g_\sigma^A$ (defined
    in Equation~\ref{eq:gA}) and the symmetrically averaged
    correlation function $g_\sigma^S$ (defined in
    Equation~\ref{eq:gS}).  The results of Monte Carlo, our
    functional, and one previously published
    functional\cite{gross2009density,
      yu2002fmt-dft-inhomogeneous-associating} are compared in each
    case.  The bottom panels show the density computed with
    Monte Carlo and with DFT.}
  \label{fig:high-density}
\end{figure*}

@@

At a higher density corresponding to a filling fraction of 0.4,
correlations are much stronger, with the bulk contact value of the
correlation function of 3.7, as seen in Figure~\ref{fig:high-density}.
This results in larger oscillations in the density at the hard
surfaces, and correspondingly more interesting behavior in the
correlation function near the interface, as shown in the bottom panels
of the plots in Figure~\ref{fig:high-density}.  The density predicted
by the White Bear functional agrees reasonably well with the
simulation results, although not so well as it did at lower density.
The discrepancies are largest in the case of the spherical cavity
(Figure~\ref{fig:high-density}d), in which the DFT considerably
underestimates the range of the density oscillations.

The asymmetric version of the correlation function (plotted
in the top panels of Figure~\ref{fig:high-density}) once again displays
relatively smooth behavior with a few small
oscillations near the interface, and a somewhat elevated value within
a diameter of the hard surface, with the magnitude of this elevation
somewhat different in each configuration.  As was the case at low
density, our correlation function $g_\sigma^A$ matches very closely
the Monte Carlo data, reproducing quite well the structure near the
interface in each configuration, although in the spherical cavity there is a small, but significant
discrepancy, comparable to the discrepancy found in the density
itself.  In each case, the correlation of Gross dramatically
underestimates the correlation at the interface, at one extreme by
40\% in the case of the spherical cavity
(Figure~\ref{fig:high-density}d), and at the other extreme by 15\%
in the test-particle scenario (Figure~\ref{fig:high-density}c).
